\documentclass[12pt]{article}

% Essential packages (basic TeX Live)
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{enumitem}
\setlist{nosep} % Global setting for compact lists
\usepackage{url}
\usepackage[alf]{abntex2cite}  % ABNT citations (Author-date)
\usepackage{float}
\usepackage{geometry}
\usepackage{times}
\usepackage{indentfirst}

% Page layout for UFPE TCC Article format
\geometry{
    a4paper,
    left=3cm,
    right=2cm,
    top=3cm,
    bottom=2cm
}

% Line spacing 1.5
\linespread{1.3}

% Settings
\graphicspath{{images/}}
\raggedbottom

% Custom title formatting for UFPE style
\makeatletter
\renewcommand{\maketitle}{%
    \begin{center}
        {\Large\bfseries\@title\par}
        \vskip 1.5em
        {\normalsize\@author\par}
        \vskip 2em
    \end{center}
}
\makeatother

% Section formatting
\usepackage{titlesec}
\titleformat{\section}{\normalfont\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalfont\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\itshape}{\thesubsubsection}{0.5em}{}

\begin{document}

\title{Quantitative Evaluation of the Circuit Breaker Pattern\\in Microservices: An Empirical Study on Resilience and Performance}

\author{
\textbf{Humberto Laff}$^{1}$, \textbf{Jamilson Ramalho Dantas}$^{1}$ (Advisor)\\[0.5em]
$^{1}$Centro de Inform\'atica (CIn)\\
Universidade Federal de Pernambuco (UFPE)\\
50740-560 -- Recife -- PE -- Brazil\\[0.5em]
\texttt{\{hlaff, jrd\}@cin.ufpe.br}
}

\maketitle

\noindent\textbf{\textit{Abstract.}} Microservices architectures have become the standard for building modern distributed systems, but synchronous communication between services introduces the risk of cascading failures. The Circuit Breaker pattern is widely recommended as a mitigation strategy, yet there is a lack of empirical studies quantifying its actual impact. This paper presents a controlled experimental evaluation comparing four service versions: V1 (Baseline without resilience), V2 (Circuit Breaker with Resilience4j), V3 (Retry with exponential backoff), and V4 (Composition of Retry + Circuit Breaker). Using Docker orchestration and k6 load testing across five realistic failure scenarios, we analyzed over 380,000 requests. Results demonstrate that Circuit Breaker improved availability from 10.3\% to 99.4\% in extreme unavailability scenarios---a gain of +89.1 percentage points. Statistical analysis confirmed a medium effect size (Cliff's Delta = 0.348, p $<$ 0.0001). Our findings provide robust empirical evidence that Circuit Breaker is essential for mission-critical synchronous microservices, while Retry alone is insufficient for persistent failures. V4 (Composition) demonstrated the highest robustness, absorbing transient jitters while maintaining protection against persistent failures.

\vspace{1em}
\noindent\textbf{Keywords:} Circuit Breaker, Microservices, Fault Tolerance, Resilience Patterns, Performance Evaluation

\vspace{2em}

% Sections
\input{sections/introducao}
\input{sections/trabalhos_relacionados}
\input{sections/metodologia}
\input{sections/resultados}
\input{sections/conclusao}

% References
\bibliographystyle{abntex2-alf}
\bibliography{references}

\end{document}
